Pangenome graph layout by Path-Guided Stochastic Gradient Descent

### 2024-02-09 manuscript/main.tex
reviewer2 q2
r2 q3

	A graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space.
	Graph layout algorithms aim to find optimal node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding of the graph.
	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetic layouts. %by ....
	This is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations alleviate such a problem \citep{Zheng2019}.
	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions.

	A \textit{pangenome} graph layout can provide a human-readable visualization of genetic variation between multiple genomes.
	However, \cite{Zheng2019}'s algorithm has a quadratic up front cost in the number of nodes to find pairwise distances to guide the layout, making it impossible to apply to pangenome graphs with millions of nodes. 
	Also, existing graph layout approaches ignore the biological information inherent in pangenome graphs.
	%Specifically, in pangenome graphs we move a single pair of nodes at a time.

@@

	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}, but the lock-free HOGWILD! method offers a highly parallelizable and thus scalable SGD approach that can be applied when the optimization problem is sparse \citep{Recht2011}.

	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to use the paths as an embedded positional system to find distances between nodes, moving pairs of nodes in parallel with a modified HOGWILD! strategy.
	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
	\red{To our knowledge, no generic graph layout algorithm takes into account such path encoded biological information when computing a graph's layout.}
	PG-SGD can be extended in any number of dimensions.
	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts.
	These algorithms have already been successfully applied to construct and visualize large-scale pangenome graphs of the Human Pangenome Reference Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.

@@

	% \FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
	The performance analysis shows that our 2D implementation outperforms \textit{BandageNG} when handling large, complex pangenome graphs.
	While \textit{BandageNG} was not able to deliver a layout of the whole HPRC graph within 1 week, our 2D PG-SGD calculated one within one day.
	%We expect that our implementation will be able to handle the next phase of the HPRC, 24 chromosomal pangenome graphs constructed from 300 individuals, without any problems.
	There are some possible optimization approaches for future work to further improve the performance of PG-SGD, making it possible for interactive use. 
	The data structure could be optimized to improve cache performance. Moreover, the high-degree of parallelism could be further exploited by using a GPU.  
	% As especially observed with chromosome 16 evaluation, our implementation is memory-bound.
	% The current data structure doesn't consider the memory access pattern.
	% Therefore, a CPU cache optimized implementation could make use of the massive parallelized calculations on a GPU.
	% Such efforts are currently in progress \url{https://github.com/nsmlzl/odgi/tree/layout_cache_optimized_cpu}.

>>>>

	A graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space.
	Graph layout algorithms aim to find optimal node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding of the graph.
	\red{One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which uses physical simulation to produce aesthetic layouts. %by ....
	This is prone to get stuck in local minima, but multi-layer strategies such as the Fast Multipole Multilevel (FM$^{3}$)~\citep{Hachul2005} method or stochastic gradient descent (SGD) implementations alleviate such a problem \citep{Zheng2019}.
	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions.}

	A \textit{pangenome} graph layout can provide a human-readable visualization of genetic variation between multiple genomes.
	However, \cite{Zheng2019}'s algorithm has a quadratic up front cost in the number of nodes to find pairwise distances to guide the layout, making it impossible to apply to pangenome graphs with millions of nodes. 
	\red{Also, existing generic graph layout approaches ignore the biological information inherent in pangenome graphs. One such bioinformatics tool is \textit{BandageNG}, the current state of the art for genome graph visualization.} 
	%Specifically, in pangenome graphs we move a single pair of nodes at a time.

@@

	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}. \red{Although, \textit{BandageNG} applies FM$^{3}$ for layout generation, it's parallelism is bound by the number of connected graph components. Alternatively, the lock-free HOGWILD! method offers a highly parallelizable and thus scalable SGD approach that can be applied when the optimization problem is sparse \citep{Recht2011}.}

	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to use the paths as an embedded positional system to find distances between nodes, moving pairs of nodes in parallel with a modified HOGWILD! strategy.
	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
	\red{To our knowledge, no generic graph layout algorithm takes into account such path encoded biological information when computing a graph's layout.}

	PG-SGD can be extended in any number of dimensions.
	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts.
	These algorithms have already been successfully applied to construct and visualize large-scale pangenome graphs of the Human Pangenome Reference Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.
	\red{In addition, we show that PG-SGD is almost an order of magnitude faster than \textit{BandageNG}.}

@@

	% \FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
	The performance analysis shows that our 2D implementation outperforms \textit{BandageNG} when handling large, complex pangenome graphs.
	While \textit{BandageNG} was not able to deliver a layout of the whole HPRC graph within 1 week, our 2D PG-SGD calculated one within one day.
	%We expect that our implementation will be able to handle the next phase of the HPRC, 24 chromosomal pangenome graphs constructed from 300 individuals, without any problems.
	There are some possible optimization approaches for future work to further improve the performance of PG-SGD, making it possible for interactive use. 
	The data structure could be optimized to improve cache performance. Moreover, the high-degree of parallelism could be further exploited by using a GPU. 
	\red{Since BandageNG was primarily designed for assembly graphs, one may have to adjust its parameters dependent on the input graph, in order to boost the layout generation or to adjust the highlighting of desired graph features.}	 
	% As especially observed with chromosome 16 evaluation, our implementation is memory-bound.
	% The current data structure doesn't consider the memory access pattern.
	% Therefore, a CPU cache optimized implementation could make use of the massive parallelized calculations on a GPU.
	% Such efforts are currently in progress \url{https://github.com/nsmlzl/odgi/tree/layout_cache_optimized_cpu}.
